ProbabilityBoundsAnalysis.jl:\\ Arithmetic with sets of distributions

### 2022-04-14 paper/paper.tex
further fixes punctuation

  \caption{Various parametrically constructed p-boxes with interval parameters.}
  \label{fig:figure3}
\end{figure*}
P-boxes generalise precise distributions and intervals in the following way. A distribution is a p-box with a precise cdf and moments, i.e. when
\begin{align*}
  \underline{F}(x) &= \overline{F}(x), \\ 
  \underline{\mu}  &= \overline{\mu}, \\ 
  \underline{\sigma}^2 &= \overline{\sigma}^2 .
\end{align*}
\noindent An interval $[a,b]$ is a p-box whose bounds are step functions
\begin{align*}
    \underline{F}(x) &= \epsilon_{b}(x) ,\\
    \overline{F}(x) &= \epsilon_{a}(x) ,
\end{align*}
\noindent where $\epsilon_k$ is
\begin{equation*}
   \epsilon_k(x) = \begin{cases} 0 &\text{when } x < k \\ 1 &\text{when } x \geq k. \end{cases} 
\end{equation*}
\noindent Moreover, theoretical bounds on the mean and variance of an interval can be found~\cite{ferson2002ramas}
\begin{align*}
  \mu &\in [a, b], \\
  \sigma^2 &\in [0, (b - a)^{2}/4].
\end{align*}
\noindent That is, it is not possible to find a distribution inside $[a, b]$ with a variance greater than $(b-a)^{2}/4$. The lower bound on the variance is zero, since scalars are also included in the interval. Under this definition of an interval, a random sample will always be the interval $[a, b]$, the cdf returns 0 when $x < a$, the interval $[0,1]$ when $a \leq x < b$, and $1$ when $b \leq x$. Further, the probability measure of the interval $X = [a, b]$ will be: 
\begin{equation*}
  \mathbb{P}_{X}(U) = \begin{cases}
    0 & \text{when } U \cap [a,b] = \emptyset \\
    1 & \text{when } [a,b] \subseteq U  \\
    [0, 1] & \text{otherwise.}\\
  \end{cases}
\end{equation*}
\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If $X$ is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step functions. Generally p-boxes may yield other probability intervals.
\subsection{Outer representations of p-boxes}
\label{sec:outer_approx}
An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the probability density, usually with quadrature. However, even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore, we require a representation of these continuous functions which is not dependent on the distribution family (for example as used in polynomial chaos expansion), and ideally is robust in the sense that it can produce an interval error for the numerical representation. Such a representation of a probability is sometimes called an "outer approximation", in the same fashion an interval is an outer approximation of a floating point number. Such outer approximations contract to the exact value or function as more computational resources are used.

@@

These p-boxes are constructed using the Chebyshev (mean and variance)~\cite{chebyshev1874valeurs}, Markov (mean and one bound)~\cite{markoff1900question} and Cantelli (mean and range) inequalities from probability theory. These inequalities allow for the cdf to be bounded from various range and moment information. For example, the Chebyshev inequality can be used to find p-box bounds from the random variables mean and variance
\begin{equation*}
  \underline{F}_{X}(x) = \begin{cases} 1 / (1 + (x - \mu_{X})^2 / \sigma^{2}_{X}),& \text{if } \mu_{X}<x \\
    1,& \text{if } x\leq \mu_{X} \end{cases}
\end{equation*}
\begin{equation*}
  \overline{F}_{X}(x)  = \begin{cases} 1 / (1 + \sigma^{2}_{X}/(x - \mu_{X})^2),& \text{if } x<\mu_{X} \\
  1,& \text{if } \mu_{X}\leq x \end{cases}
\end{equation*}
\noindent A list of available distribution-free constructors are
\begin{itemize}
  \item \textbf{MeanVar (or MeanStd)}: using Chebyshev,
  \item \textbf{MeanMin (or MeanMax)}: using Markov,
  \item \textbf{MeanMinMax}: using Cantelli,
  \item \textbf{MinMaxMeanVar (or MinMaxMeanStd)}: using~\cite{ferson2021distribution}
\end{itemize}
\noindent All distribution-free constructors may also take interval moments parameters.
\subsubsection{Operations involving intervals and distributions}
An arithmetic operation involving an interval and a precise distribution will produce set of distributions, i.e. a unique distribution for every real number in the interval, whose p-box bounds can be found. The following function performs a binary operation between a p-box and an interval
\begin{lstlisting}[language = Julia]
  using MomentArithmetic

@@

A bivariate copula (2-copula) $C$ is any function $C:[0,1]^2 \rightarrow [0,  1]$ with the following properties
\begin{enumerate}
  \item Grounded: $C(0,v) = C(u,0) = 0$,
  \item Uniform margins: $C(u,1) = u; \;C(1,v) = v$,
  \item 2-increasing: \\ $C(u_{2},v_{2}) - C_{}(u_{2}, v_{1}) - C(u_{1}, v_{2}) + C(u_{1}, v_{1}) \ge 0$\\ for all $0 \le u_{1} \le u_{2} \le 1$ and $0 \le v_{1} \le v_{2} \le 1$ .
\end{enumerate}
\noindent Three important 2-copulas are
\begin{align*}
  W(u,v) &= \mathrm{max}( u + v-1,0),  \\
  \Pi(u,v) &= uv, \\
  M(u,v) &= \mathrm{min}(u,v),
\end{align*}
\begin{figure*}[t!]
  \centering
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/biv_cdf.pdf}\hfill
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/samples.pdf}
  \caption{Cdf bounds of a bivariate p-box (right) with $X \sim \text{Beta}([4,8], 3)$, $Y \sim \text{N}([0,2], 2)$, and $\rho_{XY} = -0.8$, and 100 samples (left).}
  \label{fig:figure7}
\end{figure*}
\noindent where $W$ encodes maximally negative correlation, $\Pi$ encodes stochastic independence, and $M$ encodes maximally positive correlation. Moreover, $W$ and $M$ are bounds on all 2-copulas: $W(u,v) \leq C(u,v) \leq M(u,v)$. A bivariate cdf $F_{XY}$ can be constructed from two marginal cdfs $F_{X}$ and $F_{Y}$ and copula $C$ using Sklar's theorem
\begin{equation*}
  F_{XY}(x,y) = C(F_{X}(x), F_{Y}(y)),
\end{equation*}
\noindent where the constructed bivariate distribution will follow the dependence encoded in the copula and have the prescribed marginals. Sklar's theorem also works in reverse, where the dependence can be extracted from a multivariate distribution: 
\begin{equation*}
  C(u,v) = F_{XY}(F^{-1}_{X}(u), F^{-1}_{Y}(v)).
\end{equation*}
\noindent Copulas also give a very convenient way to sample from $F_{XY}$ in a generalisation of inverse transform sampling, where first the copula is sampled $(u,v) \sim C$, and then transformed through the inverses of the marginals $(F^{-1}_{X}(u), F^{-1}_{Y}(v))$.

>>>>

  \caption{Various parametrically constructed p-boxes with interval parameters.}
  \label{fig:figure3}
\end{figure*}
P-boxes generalise precise distributions and intervals in the following way. A distribution is a p-box with a precise cdf and moments, i.e. when
\begin{align*}
  \underline{F}(x) &= \overline{F}(x), \\ 
  \underline{\mu}  &= \overline{\mu}, \\ 
  \underline{\sigma}^2 &= \overline{\sigma}^2 .
\end{align*}
\noindent An interval $[a,b]$ is a p-box whose bounds are step functions
\begin{align*}
    \underline{F}(x) &= \epsilon_{b}(x) ,\\
    \overline{F}(x) &= \epsilon_{a}(x) ,
\end{align*}
\noindent where $\epsilon_k$ is
\begin{equation*}
   \epsilon_k(x) = \begin{cases} 0 &\text{when } x < k, \\ 1 &\text{when } x \geq k. \end{cases} 
\end{equation*}
\noindent Moreover, theoretical bounds on the mean and variance of an interval can be found~\cite{ferson2002ramas}
\begin{align*}
  \mu &\in [a, b], \\
  \sigma^2 &\in [0, (b - a)^{2}/4].
\end{align*}
\noindent That is, it is not possible to find a distribution inside $[a, b]$ with a variance greater than $(b-a)^{2}/4$. The lower bound on the variance is zero, since scalars are also included in the interval. Under this definition of an interval, a random sample will always be the interval $[a, b]$, the cdf returns 0 when $x < a$, the interval $[0,1]$ when $a \leq x < b$, and $1$ when $b \leq x$. Further, the probability measure of the interval $X = [a, b]$ is
\begin{equation*}
  \mathbb{P}_{X}(U) = \begin{cases}
    0 & \text{when } U \cap [a,b] = \emptyset, \\
    1 & \text{when } [a,b] \subseteq U,  \\
    [0, 1] & \text{otherwise.}\\
  \end{cases}
\end{equation*}
\noindent That is, if the set $U$ does not intersect the interval $X$, $\mathbb{P}_{X}=0$, i.e. $U$ certainly does not contain any of the random variables. If $X$ is fully contained in $U$, $\mathbb{P}_{X}=1$, i.e. $U$ certainly contains all of the random variables. And finally if $U$ intersects (but does not contain) $X$, the probability measure is the vacuous probability interval $\mathbb{P}_{X} = [0, 1]$, i.e. we are completely uncertain about the containment. Note that the $P_{X} = \{0, 1, [0,1]\}$ is due to the interval bounds being step functions. Generally p-boxes may yield other probability intervals.
\subsection{Outer representations of p-boxes}
\label{sec:outer_approx}
An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the probability density, usually with quadrature. However, even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore, we require a representation of these continuous functions which is not dependent on the distribution family (for example as used in polynomial chaos expansion), and ideally is robust in the sense that it can produce an interval error for the numerical representation. Such a representation of a probability is sometimes called an "outer approximation", in the same fashion an interval is an outer approximation of a floating point number. Such outer approximations contract to the exact value or function as more computational resources are used.

@@

These p-boxes are constructed using the Chebyshev (mean and variance)~\cite{chebyshev1874valeurs}, Markov (mean and one bound)~\cite{markoff1900question} and Cantelli (mean and range) inequalities from probability theory. These inequalities allow for the cdf to be bounded from various range and moment information. For example, the Chebyshev inequality can be used to find p-box bounds from the random variables mean and variance
\begin{equation*}
  \underline{F}_{X}(x) = \begin{cases} 1 / (1 + (x - \mu_{X})^2 / \sigma^{2}_{X}) & \text{if } \mu_{X}<x, \\
    1 & \text{if } x\leq \mu_{X}, \end{cases}
\end{equation*}
\begin{equation*}
  \overline{F}_{X}(x)  = \begin{cases} 1 / (1 + \sigma^{2}_{X}/(x - \mu_{X})^2) & \text{if } x<\mu_{X} \\
  1 & \text{if } \mu_{X}\leq x. \end{cases}
\end{equation*}
\noindent A list of available distribution-free constructors are
\begin{itemize}
  \item \textbf{MeanVar (or MeanStd)}: using Chebyshev,
  \item \textbf{MeanMin (or MeanMax)}: using Markov,
  \item \textbf{MeanMinMax}: using Cantelli,
  \item \textbf{MinMaxMeanVar (or MinMaxMeanStd)}: using~\cite{ferson2021distribution}.
\end{itemize}
\noindent All distribution-free constructors may also take interval moments parameters.
\subsubsection{Operations involving intervals and distributions}
An arithmetic operation involving an interval and a precise distribution will produce set of distributions, i.e. a unique distribution for every real number in the interval, whose p-box bounds can be found. The following function performs a binary operation between a p-box and an interval
\begin{lstlisting}[language = Julia]
  using MomentArithmetic

@@

A bivariate copula (2-copula) $C$ is any function $C:[0,1]^2 \rightarrow [0,  1]$ with the following properties
\begin{enumerate}
  \item Grounded: $C(0,v) = C(u,0) = 0$,
  \item Uniform margins: $C(u,1) = u; \;C(1,v) = v$,
  \item 2-increasing: \\ $C(u_{2},v_{2}) - C_{}(u_{2}, v_{1}) - C(u_{1}, v_{2}) + C(u_{1}, v_{1}) \ge 0$\\ for all $0 \le u_{1} \le u_{2} \le 1$ and $0 \le v_{1} \le v_{2} \le 1$ .
\end{enumerate}
\noindent Three important 2-copulas are
\begin{align*}
  W(u,v) &= \mathrm{max}( u + v-1,0),  \\
  \Pi(u,v) &= uv, \\
  M(u,v) &= \mathrm{min}(u,v),
\end{align*}
\noindent where $W$ encodes maximally negative correlation, $\Pi$ encodes stochastic independence, and $M$ encodes maximally positive correlation. Moreover, $W$ and $M$ are bounds on all 2-copulas: $W(u,v) \leq C(u,v) \leq M(u,v)$. A bivariate cdf $F_{XY}$ can be constructed from two marginal cdfs $F_{X}$ and $F_{Y}$ and copula $C$ using Sklar's theorem
\begin{equation*}
  F_{XY}(x,y) = C(F_{X}(x), F_{Y}(y)),
\end{equation*}
\noindent where the constructed bivariate distribution will follow the dependence encoded in the copula and have the prescribed marginals. Sklar's theorem also works in reverse, where the dependence can be extracted from a multivariate distribution: 
\begin{equation*}
  C(u,v) = F_{XY}(F^{-1}_{X}(u), F^{-1}_{Y}(v)).
\end{equation*}
\noindent Copulas also give a very convenient way to sample from $F_{XY}$ in a generalisation of inverse transform sampling, where first the copula is sampled $(u,v) \sim C$, and then transformed through the inverses of the marginals $(F^{-1}_{X}(u), F^{-1}_{Y}(v))$.
\begin{figure*}[t!]
  \centering
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/biv_cdf.pdf}\hfill
  \includegraphics[width=0.45\textwidth]{../examples/JuliaCon/fig7/samples.pdf}
  \caption{Cdf bounds of a bivariate p-box (right) with $X \sim \text{Beta}([4,8], 3)$, $Y \sim \text{N}([0,2], 2)$, and $\rho_{XY} = -0.8$, and 100 samples (left).}
  \label{fig:figure7}
\end{figure*}
